## The Effect of Weather on People and Property 
#### A study on weather in the USA from 1950 to 2011
##### By: Fred Stakem
##### Date: 8.20.14

<br />

### Synopsis

<br />

### Data Processing

Set the parameters.
```{r}
debug <- TRUE
working_dir <- '~/projects/CourseraDataScience/projects/reproducible_research/project_2/src/'
```

Setup the environment.
```{r}
setwd(working_dir)
library(R.utils)
```

Get the data and unzip it.
```{r}
data_url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
download_path <- '../raw_data/storm.csv.bz2'
data_file <- '../raw_data/debug_storm.csv'

if(!file.exists(data_file))
{
    if(!file.exists(download_path))
    {
        download.file(data_url, download_path, method='curl')
    }
    
    bunzip2(download_path, remove=FALSE)
}
```

Read the data into a data frame.
```{r}
raw_data <- read.table(data_file, fill=TRUE, sep=',', header=TRUE)
```

Question 1
*Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?*

```{r}

```
Question 2
*Across the United States, which types of events have the greatest economic consequences?*
<br />

### Results




